---
title: "session06 - homework"
format: html
editor: visual
date: 2025-12-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_libraries
#| echo: false
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesrules)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

We will be analyzing the data from the following article:

Winter, B., & Matlock, T. (2013). Making judgments based on similarity and proximity. Metaphor and Symbol, 28(4), 219-232.

```{r}
#| label: preprocessing

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv")

sim
```

In this study, they investigated the conceptual structure of the metaphor "SIMILARITY IS PROXIMITY." They did a series of experiments in which they investigated whether participants judge entities to be closer to each other when they are thought to be more similar. They presented participants with descriptions of two cities. These cities either sounded similarly or they sounded differently (`Phon`). Moreover, the descriptions depicted the cities as either semantically similar or different (`Sem`). After reading both descriptions and seeing both names, participant had to draw into a map and had to randomly guess where these two cities are located. The authors measured the `Distance` between the two cities for each participant in millimeters.

# (A) Descriptive statistics and data visualization

Let's visualize the raw `Distance` distribution for both `Sem` categories making a density plot of it and describe the distributions.

```{r}
#| label: show_overall_distribution

## Hi! Just a quick note at the beginning. When trying to commit to github, all the changes I had made somehow dissappeared. I have tried to add everything back, but I'm not really sure how good it is at this point. So sorry about that in advance. 

sim %>%
  ggplot(aes(x = Distance)) +
  geom_density(fill = 'darkseagreen') +
  theme_minimal()

# Most values seem to be between around 50 and 100. It seems like some have placed the two cities almost right next to each other, around 0 in distance, and there is a slight increase in placements between 150 and 175. I am unaware of the maximum possible distance, but it seems like few of the participants placed the cities as far away from each other as possible. The peak is taller than I had expected, which indicates that a larger portion of the reponses map a distance between 25 and 100. What we see in this plot can be a result of how the map looks, or show other things affecting the data (for instance assumptions made by participants that the map provided would be centered around the cities, real-world knowledge that cities often can be located near water, etc.) 


```

# (B) Think about priors

The researchers are interested in whether semantic similarity between the cities affects how close/far apart participants draw the cities, which we can model with the following formula:

```{r}
#| label: model_formula

mdl_formula = bf(Distance ~ Sem)

```

Use the `get_prior()` function and add `mdl_formula` as its first argument and the `sim` dataset as its second. This will give you a list of parameters that the model will estimate and will need a prior for.

```{r}
#| label: get_prior

get_prior(mdl_formula, sim) 

```

Generally, every coefficient of our model corresponds to one parameter under the class `b`, except for the `Intercept` which is its own class. So we have a prior for the `Intercept` of the linear model and a prior for the coefficient `SemSimilar`. Priors of class `b` for our predictors always receive a flat uniform prior by default if you do not specify it otherwise. The default prior for the `Intercept` is data driven and uses a student-t distributions that covers most of the data. We also have a prior for `sigma` which is the residual variance and again is by default assigned a student-t prior informed by the variance in the data.

Now I want you to specify normally distributed priors which respect the following knowledge we have: Participants needed to draw the cities on a map on a piece of din-A4 paper. The furthest two points on the map where 214 mm apart.

Your task: First use the `plot_normal()` function to find a normal distribution that covers reasonable Distance values under the above described assumptions of the experiment.

Ignore for now that you might generate values that are impossible at the tails of the distribution.

```{r}
#| label: prior_exploration
#one verion where I use the values from get_prior
plot_normal(mean = 74, sd = 3) +
  xlim(0,214)

#Trying one verion where I calculate the mean and sd
sim |> 
  summarize(M = mean(Distance, na.rm = TRUE),
            SD = sd(Distance,na.rm = TRUE),
            min = min(Distance,na.rm = TRUE),
            max = max(Distance,na.rm = TRUE))

plot_normal(mean = 77.6, sd = 40.1) +
  xlim(0,214)

```

# (C) Specify priors

Now based on your exploration, specify priors. Remember what the two coefficients `Intercept` and `SemSimilar` mean: `Intercept` reflects the average Distance measure for `Sem = Different`, and `SemSimilar` reflects the estimated difference between the Intercept and `Sem = Similar`. Define a prior that assumes normally distributed Distance measures somewhere in the middle of the map for both conditions:

```{r}
#| label: spec_priors

my_priors <- c(prior(normal(77.6, 40.1), class = 'Intercept'),
                  prior(normal(77.6, 40.1), class = 'b'),
                   prior(normal(0, 10), class = 'sigma'))


my_priors

```

# (D) Run a prior predictive check

Add your priors to the model below. We will run the model and sample only from our priors, i.e. we ignore the data. This is a good sanity check to see if our priors make sense and what they would predict. Plug in `my_priors` from (C) and run the code chunk.

```{r}
#| label: model_priors

xmdl_prior <- brm(Distance ~ Sem,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            data = sim,
            seed = 42)


# We do a little wrangling to get it into a nice shape
xmdl_prior |> 
  # first extract the samples
  spread_draws(b_Intercept, b_SemSimilar) |> 
  # Create two columns that represent the two SEM categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_SemSimilar) |> 
  # and then bring into a long format like the data
  select(Different, Similar) |> 
  pivot_longer(everything(), names_to = "Sem", values_to = "Distance") |> 
  # now we plot
  ggplot(aes(x = Distance, fill = Sem)) +
    geom_density() +
    facet_wrap(Sem ~ .)

```

Do the priors reflect what you had in mind? If not, go back to (C) and play around with the priors until you are satisfied.

# (E) Run the full model

If you are happy with your priors, run the full model and plot the posteriors just like in the code chunk above. How did the posteriors change after seeing the data?

```{r}
#| label: model_posteriors

xmdl <- brm(Distance ~ Sem,
            family = gaussian(),
            data = sim,
            seed = 42)
summary(xmdl)

#plotting posteriors

xmdl |> 
  # first extract the samples
  spread_draws(b_Intercept, b_SemSimilar) |> 
  # Create two columns that represent the two SEM categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_SemSimilar) |> 
  # and then bring into a long format like the data
  select(Different, Similar) |> 
  pivot_longer(everything(), names_to = "Sem", values_to = "Distance") |> 
  # now we plot
  ggplot(aes(x = Distance, fill = Sem)) +
    geom_density() +
    facet_wrap(Sem ~ .)

#The posteriors look a bit more "neat" than the actual data. The actual data also show steeper slopes than the posteriors. They also look different now, because I don't remember what I finally set for my priors, which makes the posteriors having a large variance in distance, between -1000 and 1000, which obviously isn't ideal. 

```

After seeing the data the very diffuse priors have given way to much more narrow and shifted posteriors.

We can also plot the data in an alternative way, since we are mostly interested in the difference between the two conditions:

```{r}
#| label: alternative_plot

# We do a little wrangling to get it into a nice shape
xmdl |> 
  # first extract the samples for the differences only
  spread_draws(b_SemSimilar) |> 
  # plot
  ggplot(aes(x = b_SemSimilar)) +
    geom_density(fill = "purple") 

```

Now, one last step, before we move on. Let's check whether this model could have generated the data using `pp_check()`.

Throughout our modeling process, we'll constantly use `pp_check()`, which is a way for us to know whether our model could've generated the data. This is a way of investigating whether the model 'fits' well. PP checks can tell us, for example, whether the chosen response distribution (in this case Gaussian) was appropriate, or whether the model will have to be expanded, that is, whether particular predictors are missing.

Describe whether the data (dark line) and the posterior samples (light blue line) match up and if now, what the nature of the mismatch is.

```{r}
#| label: pp_check

pp_check(xmdl, ndraws = 100)

# The data and posteriors are not a complete match. The actual data has a more "jagged" curve, with more variance between 100 and 200, and is steeper than the posterior samples. I don't know how good a fit it needs to be, but it seems like the posterior samples does not quite represent the density of the data at around 75, and shows more even slopes.


```

# OPTIONAL (F) fix the assumption violation

Can you fix the mismatch between the observed data (dark blue line) and the simulated data (light blue lines) with, for example, a transformation of the Distance measure? Run the model again with the transformed values and do the pp check again.

```{r}
#| label: transform

# YOUR CODE HERE

```

# OPTIONAL (G) specify priors that will reduce the effect of SEM to zero

Now I want you to specify priors that will dominate the posterior, i.e. they will eliminate any predicted difference between conditions.

```{r}
#| label: too-strict_priors

# YOUR CODE HERE

```
